Compositional Certified Resource Bounds

### 2015-04-15 paper/main.tex
move renewcommand{\max} before Table 2 caption
pass through derivation Rules

The rules of our derivation system must serve two purposes.  They must
\emph{attach} potential to certain program variable intervals and use
this potential, when it is allowed, to \emph{pay} for resource consuming
operations.  These two purposes are illustrated on the {\sc Q:Skip}
rule.  This rule reuses its precondition as postcondition, it is
explained by two facts: first, no resource is consumed by the skip
operation, thus no potential has to be used to pay for the evaluation;
second, the program state is not changed by
the execution of a skip statement, thus all the potential available
before the execution of the skip statement is still available after.
\iffull{
These two observations would also justify a rule with the side-condition
$Q \ge Q'$, because some potential is always safely lost.  For the sake
of orthogonality, it is not how we present the rule and this form of
weakening is in fact expressed as the stand-alone rule {\sc Q:Weak}
described later.}{}

@@

The rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc Q:Inc}
describe how the potential is distributed after a size change of
a variable.
The rule {\sc Q:IncP} is for increments $x \gets x + y$
and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
that can be applied if $y$ is not positive.
The rules are all equivalent in the case where $y=0$.
The rule {\sc Q:Inc} can be applied
if we cannot find the sign of $y$.  % In $\toolname$,
% the rules for increments and decrements are combined into one rule\iffull{ where
% the specialized rules take priority over {\sc Q:Inc}.}{.}

@@

The rule {\sc Q:Loop} is a cornerstone of our analysis.  To apply it on a
loop body one needs to find an invariant potential $Q$ that will pay for
the iterations.  At each iteration, $M_l$ resources are spent to jump
back, this explains the postcondition $Q{+}M_l$.  To exit the loop,
the programmer uses the \code{break} statement, it is the reason why
the postcondition for the full $\code{loop}~S$ statement $\{ \Gamma';
Q' \}$ is used as break postcondition in the derivation for $S$.

Another interesting rule is {\sc Q:Call}.  It needs to account for the
changes to the stack caused by the function call, the arguments/return
value passing, and the preservation of local variables.  We can sum up
the main ideas of the rule as follows.
\itemskip
\begin{itemize}
\item The potential in the pre- and postcondition of the function
  specification is equalized to its matching potential in the callee's
  pre- and postcondition.
\itemskipIn
\item
  The potential of intervals $|\inter x y|$ is preserved
  across a function call
  if $x$ and $y$ are local\iffull{ variables.}{.}
\item
  The unknown potentials after the call (e.g.
  $|\inter x g|$, with $x$ local and $g$ global)
  are set to zero in the postcondition.
\end{itemize}
\itemskip
If $x$ and $y$ are local variables and $f(x,y)$ is
called, {\sc Q:Call} splits the potential of
$|\inter x y|$ in two parts, one part to perform the
computation in the function $f$ and one part to keep for
later use after the function call.  This splitting is
realized by the equations $Q = P {+} S$ and $Q' =
P' {+} S'$.  Arguments in the function precondition
$(\Gamma_f; Q_f)$ are named using a fixed vector $\Vargs$
of names different from all program variables.  This
prevents name conflicts from happening and ensures that
the substitution $[\Vargs/\vec x]$ is meaningful.
Symmetrically, we use the unique name $\Vret$ to represent
the return value in the function's postcondition
$(\Gamma'_f; Q'_f)$.

The rule {\sc Q:Weak} is the only rule not syntax directed. %In the implementation
%we apply {\sc Q:Weak} before loops and between the two statements of a
%sequential composition.
We could integrate weakenings into every syntax directed rule but, for
the sake of efficiency, the implementation uses a simple heuristic instead.
The high-level idea of {\sc Q:Weak} is the following: If
we have a sound judgement, then it is sound to add more potential to the
precondition and remove potential from the postcondition.  The concept of
\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
that is defined in the rule {\sc Relax}. This rule
also deals with the important task of transferring constant potential
(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
from the logical context that the interval size $|\inter x y| \geq l$ is
larger than a constant $l$ then we can turn the potential $q_{xy} {\cdot}
|\inter x y|$ form the interval into the constant potential $l{\cdot}q_{xy}$
and guarantee that we do not gain potential.  Conversely, if $|\inter
x y| \leq u$ for a constant $u$ then we can transfer constant potential
$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
without gaining potential.

@@

\begin{table*}
\nocaptionrule \caption{
  Experimental evaluation comparing the bounds generated KoAT, Rank,
  LOOPUS, SPEED, and our automatic amortized analysis on several
  challenging linear examples. Results for KoAT and SPEED were extracted
  from previous publications because KoAT cannot take C programs as
  input in its current version and SPEED is not available.
  Entries marked with ? indicate that we cannot test the respective example with the tool. Entries
  marked with --- indicate that the tool failed to produce a result. We write
  $\max(a,b)$ for the maximum of $a$ and $b$.
}
\vspace{6pt}
\label{tab:eval}
\footnotesize

\newcommand{\dir}[2]{\texttt{#1} \\ \texttt{#2}}
\newcommand{\file}[2][c]{%
  \begin{tabular}[#1]{@{}r@{}}#2\end{tabular}}
\renewcommand{\max}[0]{{\rm mx}}

>>>>

The rules of our derivation system must serve two purposes.  They must
\emph{attach} potential to certain program variable intervals and use
this potential, when it is allowed, to \emph{pay} for resource consuming
operations.  These two purposes are illustrated on the {\sc Q:Skip}
rule.  This rule reuses its precondition as postcondition, it is
explained by two facts: First, no resource is consumed by the skip
operation, thus no potential has to be used to pay for the evaluation.
Second, the program state is not changed by
the execution of a skip statement. Thus all potential available
before the execution of the skip statement is still available after.
\iffull{
These two observations would also justify a rule with the side-condition
$Q \ge Q'$, because some potential is always safely lost.  For the sake
of orthogonality, it is not how we present the rule and this form of
weakening is in fact expressed as the stand-alone rule {\sc Q:Weak}
described later.}{}

@@

The rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc Q:Inc}
describe how the potential is distributed after a size change of
a variable.
The rule {\sc Q:IncP} is for increments $x \gets x + y$
and {\sc Q:DecP} is for decrements $x \gets x - y$. They both apply only
when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
Of course, there are symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
(not presented here) that can be applied if $y$ is negative.
The rules are all equivalent in the case where $y=0$.
The rule {\sc Q:Inc} can be applied
if we cannot find the sign of $y$.  % In $\toolname$,
% the rules for increments and decrements are combined into one rule\iffull{ where
% the specialized rules take priority over {\sc Q:Inc}.}{.}

@@

The rule {\sc Q:Loop} is a cornerstone of our analysis.  To apply it
on a loop body, one needs to find an invariant potential $Q$ that will
pay for the iterations.  At each iteration, $M_l$ resources are spent
to jump back. This explains the postcondition $Q{+}M_l$. Since the
loop can only be exited with a \code{break} statement, the
postcondition $\{ \Gamma'; Q' \}$ for the statement $\code{loop}~S$ is
used as break postcondition in the derivation for $S$.

Another interesting rule is {\sc Q:Call}.  It needs to account for the
changes to the stack caused by the function call, the arguments/return
value passing, and the preservation of local variables.  We can sum up
the main ideas of the rule as follows.
\itemskip
\begin{itemize}
\item The potential in the pre- and postcondition of the function
  specification is equalized to its matching potential in the callee's
  pre- and postcondition.
\itemskipIn
\item
  The potential of intervals $|\inter x y|$ is preserved
  across a function call
  if $x$ and $y$ are local\iffull{ variables.}{.}
\item
  The unknown potentials after the call (e.g.
  $|\inter x g|$, with $x$ local and $g$ global)
  are set to zero in the postcondition.
\end{itemize}
\itemskip
If $x$ and $y$ are local variables and $f(x,y)$ is
called, {\sc Q:Call} splits the potential of
$|\inter x y|$ in two parts. One part to perform the
computation in the function $f$ and one part to keep for
later use after the function call.  This splitting is
realized by the equations $Q = P {+} S$ and $Q' =
P' {+} S'$.  Arguments in the function precondition
$(\Gamma_f; Q_f)$ are named using a fixed vector $\Vargs$
of names different from all program variables.  This
prevents name conflicts and ensures that
the substitution $[\Vargs/\vec x]$ is meaningful.
Symmetrically, we use the unique name $\Vret$ to represent
the return value in the function's postcondition
$(\Gamma'_f; Q'_f)$.

The rule {\sc Q:Weak} is the only rule that is not syntax directed. %In the implementation
%we apply {\sc Q:Weak} before loops and between the two statements of a
%sequential composition.
We could integrate weakenings into every syntax directed rule but, for
the sake of efficiency, the implementation uses a simple heuristic instead.
The high-level idea of {\sc Q:Weak} is the following: If
we have a sound judgement, then it is sound to add more potential to the
precondition and remove potential from the postcondition.  The concept of
\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
that is defined in the rule {\sc Relax}. This rule
also deals with the important task of transferring constant potential
(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
from the logical context that the interval size $|\inter x y| \geq \ell$ is
larger than a constant $\ell$ then we can turn the potential $q_{xy} {\cdot}
|\inter x y|$ form the interval into the constant potential $\ell{\cdot}q_{xy}$
and guarantee that we do not gain potential.  Conversely, if $|\inter
x y| \leq u$ for a constant $u$ then we can transfer constant potential
$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
without gaining potential.

@@

\begin{table*}
\renewcommand{\max}[0]{{\rm mx}}
\nocaptionrule \caption{
  Experimental evaluation comparing the bounds generated KoAT, Rank,
  LOOPUS, SPEED, and our automatic amortized analysis on several
  challenging linear examples. Results for KoAT and SPEED were extracted
  from previous publications because KoAT cannot take C programs as
  input in its current version and SPEED is not available.
  Entries marked with ? indicate that we cannot test the respective example with the tool. Entries
  marked with --- indicate that the tool failed to produce a result. We write
  $\max(a,b)$ for the maximum of $a$ and $b$.
}
\vspace{6pt}
\label{tab:eval}
\footnotesize

\newcommand{\dir}[2]{\texttt{#1} \\ \texttt{#2}}
\newcommand{\file}[2][c]{%
  \begin{tabular}[#1]{@{}r@{}}#2\end{tabular}}
